Deployable Probabilistic Programming

### 2019-04-17 paper/paper.tex
fixed bibliography to look reasonably well

Probabilistic programming is traditionally associated with
design and implementation of probabilistic programming
languages. An apparent justification for a new programming
language is that a probabilistic program has a different
semantics~\cite{GHNR14} than a `regular' program of similar
structure. Goodman and Stuhlm\"uller~\cite{GS15} offer a
systematic approach to design and implementation of
probabilistic programming languages based on extension of the
syntax of a (subset of) regular programming language and
transformational compilation to continuation-passing form. Van
de Meent et al~\cite{MPY+18} give a comprehensive overview of
available choices of design and implementation of first-order
and higher-order probabilistic programming languages.
Stan~\cite{Stan17} is built around an imperative probabilistic
programming language with program structure tuned for most
efficient execution of inference on the model.
SlicStan~\cite{GGS19} is built on top of Stan as a
source-to-source compiler, providing the model developer with
a more intuitive language while retaining performance benefits
of Stan. Our work differs from existing approaches in that we
advocate enabling probabilistic programming in a general purpose
programming language by leveraging existing capabilities of the
language, rather than building a new language on top or besides
an existing one.

The need for integration between probabilistic programs and the
surrounding software environment is well understood in the
literature~\cite{TMY+16,Pyro18}. Probabilistic programming
languages are often implemented as embedded domain-specific
languages~\cite{SGG15,GXG18,TMY+16} and include a mechanism of calling
functions from libraries of the host languages. Our work goes a
step further in this direction: since the language of
implementation of probabilistic models coincides with the host
language, any library or user-defined function of the host
language can be directly called from the probabilistic model,
and model methods can be directly called from the host language.

Differentiable programming for machine learning (fundamentals,
Stan, Tangent, Flux, Zygote, Pyro, Edward).

@@

\begin{lstlisting}
hmc := &infer.HMC{
	Eps: 0.1,
}
samples := make(chan []float64)
hmc.Sample(m, x, samples)
for i := 0; i != 1000; i++ {
	x = <-samples
}
hmc.Stop()
\end{lstlisting}

>>>>

Probabilistic programming is traditionally associated with
design and implementation of probabilistic programming
languages. An apparent justification for a new programming
language is that a probabilistic program has a different
semantics~\cite{GHNR14} than a `regular' program of similar
structure. Goodman and Stuhlm\"uller~\cite{GS15} offer a
systematic approach to design and implementation of
probabilistic programming languages based on extension of the
syntax of a (subset of) regular programming language and
transformational compilation to continuation-passing form. Van
de Meent et al~\cite{MPY+18} give a comprehensive overview of
available choices of design and implementation of first-order
and higher-order probabilistic programming languages.
Stan~\cite{Stan17} is built around an imperative probabilistic
programming language with program structure tuned for most
efficient execution of inference on the model.
SlicStan~\cite{GGS19} is built on top of Stan as a
source-to-source compiler, providing the model developer with
a more intuitive language while retaining performance benefits
of Stan. Our work differs from existing approaches in that we
advocate enabling probabilistic programming in a general 
programming language by leveraging existing capabilities of the
language, rather than building a new language on top or besides
an existing one.

The need for integration between probabilistic programs and the
surrounding software environment is well understood in the
literature~\cite{TMY+16,BCJ+19}. Probabilistic programming
languages are often implemented as embedded domain-specific
languages~\cite{SGG15,GXG18,TMY+16} and include a mechanism of calling
functions from libraries of the host languages. Our work goes a
step further in this direction: since the language of
implementation of probabilistic models coincides with the host
language, any library or user-defined function of the host
language can be directly called from the probabilistic model,
and model methods can be directly called from the host language.

Automatic differentiation is widely employed in machine
learning~\cite{BPR+17,MBB+18} where it is also known as
`differentiable programming', and is responsible for enabling
efficient inference in many probabilistic programming
systems~\cite{Stan17,GXG18,ISF+18,BCJ+19,THS+17}. Different
automatic differentiation techniques~\cite{GW08} allow 
different compromises between flexibility, efficiency, and
feature-richness~\cite{SDT14,PGC+17,ISF+18}.  Automatic
differentiation is usually implemented through either
operator overloading~\cite{Stan17,PCG+17,GXG18} or source
code transformation~\cite{ISF+18,MMW18}. Our work, too, relies
on automatic differentiation, implemented as source code
transformation. However, a novelty of our approach is that
instead of using explicit directives to denote parts of code
which needs to be differentiated, we rely on the type system of
Go to selectively differentiate the code relevant for inference,
thus combining advantages of both operator overloading and
source code transformation.

@@

\begin{lstlisting}
hmc := &infer.HMC{
	Eps: 0.1,
}
samples := make(chan []float64)
hmc.Sample(m, x, samples)
for i := 0; i != 5000; i++ {
	x = <-samples
}
hmc.Stop()
\end{lstlisting}
